Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-08-25 draft.tex
included results figure
Added revised introduction from Sylvia

\section{\uppercase{Introduction}}
\label{sec:Introduction}
Epidemiology is the study of causation of diseases.
%%
Large population studies, such as the Study of Health in Pomerania (SHIP) \ref{SHIP} gather as much information as possible about participants to be assessed towards different diseases.
%%
These information are used to determine risk factors for diseases, helping people to make their lifestyle healthier or helping in diagnosing a disease.
%%
Epidemiological research is strongly hypothesis driven.
%
Observations made by clinicians are translated into hypothesis, which are then statistically evaluated using data variables from epidemiological studies.

Modern cohort studies often comprise medical image data.
%%
\com{Short summary}
%%

Back pain is one of the most frequent diseases in the western civilization.
%%

Our goal is to combine data mining algorithm with data visualization to provide insight into the quality of image derived data to analyze if it acts as a risk factor for a disease.
%\com{Combine Data Mining Algo}

More than 5 variables are rarely analyzed simultaneously.
%%
%The variables analyzed simultaneously is limited to 
Our contributions are:
\begin{itemize}
	\item Analyzing back pain using image-derived variables of 2,240 subjects.
	\item Assessing the suitability of lumbar spine shape for diagnosing back pain
	\item Analyzing correlations between between image-based and socio-demographic as well as medical parameters.
	\item Identification of most important variables using data mining methods \com{noch schoen schreiben}
	\item semiquantitative Auswertmöglichkeiten werden dem User oft zur verfügung gestellt
	\item Techniken sind als Teil des IVA-Frameworkes zu verstehen
\end{itemize}

@@

%\com{ToDo: SQRT Skalierung der Treesize-Dimension, da hauptsaechlich geringe Groessen von interesse sind, ausreisser aber diesen Bereich schwer durchschaubar machen}
% %%
%
% %%
% Abstracting the classification results into two continuos variables allows us to visualize the result of 134 classifications in one scatter plot.
% %%
% By mapping the group affiliation (in our case body-size) to color or shape, we can even plot all 402 results in a single visualization (Fig.~\ref{fig:decision-tree} a).
%%
\subsubsection{Dummy Variables}
Dummy variables abstract a categorical variable with multiple manifestations into several binary variables.
%%
Each binary variables encodes the presence of a manifestation.
%%
For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four binary variables (\emph{No pain - Yes/No} to \emph{Large Pain - Yes/no}).
%%
One subject can only have one of these variables set to true.
%%
This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data parameter.
%%\com{ToDo: Dummy-Variables}
%%------------
\subsubsection{Interaction With the Visualization}
%%
The described visualization provides a good overview over the classification results.
%%
We however still want to be able to look display \emph{details-on-demand} \ref{shneiderman1996} and examine a decision tree in detail.
%%
This is realized by clicking on a entry on the visualization, which then displays the corresponding decision tree in detail (Fig.~\ref{fig:decision-tree} b).
%%
This allows for sequentially analyzing the classifications.
%%------------
\subsubsection{Implementation}
All analysis are carried out using \texttt{R}, a popular programming language for statistical calculations and visualizations.
%%
The interactive visualizations are realized using the \texttt{ggvis}\footnote{GGVIS Footnote} package.
%%
As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization parameters using UI controls, such as sliders.
%%
We implemented our application as web based prototype using \texttt{R Shiny}\footnote{R Shiny footnote}.
%%
The package combines the power of \texttt{R} with advantages of web-based applications.
%%
It allows us to quickly exchange results with our collaborating epidemiological partners.
%%
They can try out the technique without installing anything.
%%
Exchanging software becomes as easy as exchanging a hyperlink.
%%
\com{TODO: Auf Eignung von RMarkdown als Report eingehen - ergebnisse koennen schnell publiziert und mit neuen Daten durchgetestet werden.}
%%---------
\subsection{Results}
We ran the analysis using different subject groups:
\begin{enumerate}
	%%
	\item All subjects,
	%%
	\item subdivision into \emph{male} and \emph{female},
	%%
	\item subdivision into \emph{age quantiles}, yielding the groups (21, 43.8] (43.8, 54] (54, 64] (64, 90],
	%%
	\item subdivision into \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
	%%
\end{enumerate}
We plotted each group twice.
%%
The first plot shows all original variables.
%%
The second plot shows all categorical variables transformed into dichotomous dummy variables.

>>>>

\section{\uppercase{Introduction}}
\label{sec:Introduction}
Epidemiology is the study of causation of diseases. For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) [??] gather as much information as possible about participants to be assessed towards different diseases.
%%
These information are used to determine risk factors for diseases, help people to improve their lifestyle and their health or to support diagnosis of widespread diseases.
%%
Epidemiological research is strongly hypothesis-driven.
%%
More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
%%
As a result, the hypothesis will be accepted or rejected.
%%

%For epidemiological research, cohort studies are acquired.
For epidemiological research, cohort studies are carried out.
%%
These studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes like gender, age and weight.
%%
Back pain is one of the most frequent diseases in the western civilization. 
%%
Epidemiological analysis of back pain aim to characterize the disease by determining risk factors.
%Many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%Since back pain is one of the most frequent diseases in the western civilization, many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%%
Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for the presence of back pain, no automatic classification approach to characterize back pain based on lumbar back spine attributes has been established yet.
%%
Furthermore, previous analyses mainly focused on a single parameter or a small subset of parameters [Quelle].
%%

In this paper, we present a thorough analysis of the image-derived data from lumbar back spine datasets and analyze their discriminating power for different characteristics of the disease.
%%
Therefore, we combine data mining algorithms with data visualization techniques.
%%
%Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
%%
\com{Da in der Auflistung oefter vom IVA-Framework die Rede ist: Wollen wir hierzu noch was schreiben, z.B. dass das im VIS-Paper definiert wurde und was wir darunter verstehen; sowie dass diese Arbeit als Erweiterung dieses Frameworks gedacht ist?}

Our contributions are:
\begin{itemize}
\item An IVA-framework for back pain analysis based on image-derived variables of 2,240 subjects \com{Ist diskutabel, so richtig interaktiv ist es nicht. IVA wurde auch noch nicht vorher definiert.}
%%
\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis
%%
\item Detection of correlations between image-based and socio-demographic as well as medical parameters for an improved hypothesis generation
%%
\item Identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework \com{Wieder: IVA wurde noch nicht definiert}
%%
\end{itemize}

% -- Old Introduction
% Epidemiology is the study of causation of diseases.
% %%
% Large population studies, such as the Study of Health in Pomerania (SHIP) \ref{SHIP} gather as much information as possible about participants to be assessed towards different diseases.
% %%
% These information are used to determine risk factors for diseases, helping people to make their lifestyle healthier or helping in diagnosing a disease.
% %%
% Epidemiological research is strongly hypothesis driven.
% %
% Observations made by clinicians are translated into hypothesis, which are then statistically evaluated using data variables from epidemiological studies.
%
% Modern cohort studies often comprise medical image data.
% %%
% \com{Short summary}
% %%
%
% Back pain is one of the most frequent diseases in the western civilization.
% %%
%
% Our goal is to combine data mining algorithm with data visualization to provide insight into the quality of image derived data to analyze if it acts as a risk factor for a disease.
% %\com{Combine Data Mining Algo}
%
% More than 5 variables are rarely analyzed simultaneously.
% %%
% %The variables analyzed simultaneously is limited to
% Our contributions are:
% \begin{itemize}
% 	\item Analyzing back pain using image-derived variables of 2,240 subjects.
% 	\item Assessing the suitability of lumbar spine shape for diagnosing back pain
% 	\item Analyzing correlations between between image-based and socio-demographic as well as medical parameters.
% 	\item Identification of most important variables using data mining methods \com{noch schoen schreiben}
% 	\item semiquantitative Auswertmöglichkeiten werden dem User oft zur verfügung gestellt
% 	\item Techniken sind als Teil des IVA-Frameworkes zu verstehen
% \end{itemize}

@@

%\com{ToDo: SQRT Skalierung der Treesize-Dimension, da hauptsaechlich geringe Groessen von interesse sind, ausreisser aber diesen Bereich schwer durchschaubar machen}
% %%
%
% %%
% Abstracting the classification results into two continuos variables allows us to visualize the result of 134 classifications in one scatter plot.
% %%
% By mapping the group affiliation (in our case body-size) to color or shape, we can even plot all 402 results in a single visualization (Fig.~\ref{fig:decision-tree} a).
%%
\subsubsection{Dummy Variables}
Dummy variables abstract a categorical variable with multiple manifestations into several binary variables.
%%
Each binary variables encodes the presence of a manifestation.
%%
For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four binary variables (\emph{No pain - Yes/No} to \emph{Large Pain - Yes/no}).
%%
One subject can only have one of these variables set to true.
%%
This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data parameter.
%%\com{ToDo: Dummy-Variables}
%%------------
\subsubsection{Interaction With the Visualization}
%%
The described visualization provides a good overview over the classification results.
%%
We however still want to be able to look display \emph{details-on-demand} \ref{shneiderman1996} and examine a decision tree in detail.
%%
This is realized by clicking on a entry on the visualization, which then displays the corresponding decision tree in detail (Fig.~\ref{fig:decision-tree} b).
%%
This allows for sequentially analyzing the classifications.
%%------------
\subsubsection{Implementation}
All analysis are carried out using \texttt{R}, a popular programming language for statistical calculations and visualizations.
%%
The interactive visualizations are realized using the \texttt{ggvis}\footnote{GGVIS Footnote} package.
%%
As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization parameters using UI controls, such as sliders.
%%
We implemented our application as web based prototype using \texttt{R Shiny}\footnote{R Shiny footnote}.
%%
The package combines the power of \texttt{R} with advantages of web-based applications.
%%
It allows us to quickly exchange results with our collaborating epidemiological partners.
%%
They can try out the technique without installing anything.
%%
Exchanging software becomes as easy as exchanging a hyperlink.
%%
\com{TODO: Auf Eignung von RMarkdown als Report eingehen - ergebnisse koennen schnell publiziert und mit neuen Daten durchgetestet werden.}
%%---------
\subsection{Results}
\begin{figure*}[p!]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=1.0\textwidth]{figures/results}
  \caption{myResults}
  \label{fig:results}
\end{figure*}
We ran the analysis using different subject groups:
\begin{enumerate}
	%%
	\item All subjects,
	%%
	\item subdivision into \emph{male} and \emph{female},
	%%
	\item subdivision into \emph{age quantiles}, yielding the groups (21, 43.8] (43.8, 54] (54, 64] (64, 90],
	%%
	\item subdivision into \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
	%%
\end{enumerate}
We plotted each group twice.
%%
The first plot shows all original variables.
%%
The second plot shows all categorical variables transformed into dichotomous dummy variables.
